```{r} 

library(DESeq2)
library(ggplot2)
library(tidyverse)
library(GEOquery)
library(RColorBrewer)
library(org.Hs.eg.db)
library(clusterProfiler)
library(ggridges)
library(pheatmap)

```
```{r}
TDP_data <- getGEO("GSE128647")
df<- pData(phenoData(TDP_data[[1]]))

original_values <- c("GFP", "WT","FL","R151A","D247A")
replacement_values <- c("Negative control", "TDP43 wild type", "2phe_RRM1","Arg_RRM1to_Ala","Asp_RRM2_Ala")

df$source_name_ch1 <- ifelse(df$source_name_ch1 %in% original_values, replacement_values[match(df$source_name_ch1, original_values)], df$source_name_ch1)
new_df <-data.frame(Condition = df$source_name_ch1, Sample=rownames(df))
rownames(new_df)<-df$title


```

Reading the raw Data and factoring the condition feature in the metadata.
```{r}
raw_data <- read.table("GSE128647_htseq_raw_counts.txt",header=T)
colnames(raw_data) <- sub("^Sample_", "", colnames(raw_data))
rownames(raw_data) <- raw_data$id
raw_data <- raw_data[,-1]
all(colnames(raw_data) %in% rownames(new_df)) #True metadata matches raw data
all(colnames(raw_data)==rownames(new_df)) #True metadata has same order as raw data
new_df$Condition<-factor(new_df$Condition) #factoring
new_df$Condition<-relevel(new_df$Condition,ref="Negative control")
str(new_df$Condition)


```
# Preprocessing steps:

-Creating DESeq object

```{r}
dds<- DESeqDataSetFromMatrix(countData =raw_data,
                       colData = new_df,
                       design= ~ Condition)
```
- filtering out rows with low counts across all samples before normalization
```{r}

keep<-rowSums(counts(dds))>=10
dds<- dds[keep,]
```
- Running DESeq and obtaining the normalized counts based on the size factor
```{r}
dds <- DESeq(dds)


```
- Results:
Genes with p adjusted values greater than 0.05 are disregarded in the analysis
The PCA plot highlights the contrast between samples with wild type TDP-43 and samples with mutated versions of TDP-43. Moreover, samples with mutations in RRM1 are highly similar.
```{r}
MA_plot <- plotMA(dds)



vsd <- vst(dds, blind=FALSE) #obtaining the variation stabilizing transformations that transforms the variance to become more normal (allowing more spread) and independent from the mean 
#PCA
plotPCA(vsd, intgroup=c("Condition"))
```
```{r}
#HeatMap similarities between samples

sampledist <- dist(t(assay(vsd)))
sampledist_matrix <-as.matrix(sampledist)
colnames(sampledist_matrix)<- new_df$Condition
rownames(sampledist_matrix)<- new_df$Condition
colors<- colorRampPalette(rev(brewer.pal(9,"Blues")))(255)
pheatmap(sampledist_matrix,clustering_distance_rows=sampledist, clustering_distance_cols=sampledist,col=colors)

```
```{r}
library(dplyr)
library(org.Hs.eg.db)
# conditions vs TDP-43 wild type!!


all_genes <-list()
for (i in resultsNames(dds)){ #loop and get results for all conditions
  if (i != 'Intercept'){
    
    res<- results(dds,name=i,alpha=0.05)
    # shrink the result to obtain more accurate values
    res_shrunken<-lfcShrink(dds,coef=i,type='apeglm',res=res)
    res_shrunken_df<- data.frame(res_shrunken)
    
    
    
    
    sig_genes<- res_shrunken_df %>% filter(padj<=0.05, abs(log2FoldChange)>1) #filtering significant genes
    # REMOVE GENES THAT HAVE NO NAME MAPPING from the significant genes
    genes<-rownames(sig_genes) #get gene esemble id
    genes_name<-mapIds(org.Hs.eg.db, keys = genes, column = "SYMBOL", keytype = "ENSEMBL") #map them to gene names
    sig_genes$gene_names<- genes_name
    sig_genes$gene_names[sig_genes$gene_names == "<NA>"] <- NA
    sig_genes<-sig_genes[!is.na(sig_genes$gene_names), ]
    
    ###
    
    
    # Get the top 25 rows with the greatest absolute values of log2FoldChange
    top_25_genes <- sig_genes %>%
                 arrange(desc(abs(log2FoldChange))) %>%  # Arrange by absolute log2FoldChange in descending order
                top_n(25)  # Select the top 25 genes 
    pdf(file = paste0("plot_", i, ".pdf"), width = 10, height = 6)
    plotCounts(dds, gene = row.names(top_25_genes)[1], intgroup = 'Condition', normalized = TRUE) # plot the most differentiated gene in each condition against all conditions
    dev.off() #close pdf device
    all_genes[[i]]<-list(top25 =top_25_genes,ress=res_shrunken_df)
   
   
  }
  
}
```

```{r}
#Volcano plots
for (j in resultsNames(dds)){
if (j!="Intercept"){

#gene=row.names(all_genes[[j]]$res)
  
all_genes[[j]]$ress <- all_genes[[j]]$ress %>%
  mutate(diffexp = ifelse(log2FoldChange > 0, "overexpressed", "underexpressed"))%>% filter(!is.na(padj))


 volcano <-ggplot(all_genes[[j]]$ress, aes(x =log2FoldChange, y = -log10(padj),col=diffexp,label = rownames(all_genes[[j]]$ress ))) +

  theme_minimal() +
  labs(
    title = j,
    x = "Log2FoldChange",
    y = "-Log10(p-value)"
  )
plot(volcano+geom_text(check_overlap = TRUE))
volcano_ <-ggplot(all_genes[[j]]$ress, aes(x =log2FoldChange, y = -log10(padj),col=diffexp,main=j ))+
  geom_point(aes(color = ifelse(padj  < 0.05  & diffexp == "overexpressed" , "blue",ifelse(diffexp == "underexpressed" & padj < 0.05, "red", "black"))), size = 2) +
  
  
  scale_color_manual(values = c("black", "red",'blue'),
                     labels = c("Not Significant", "Overexpressed (p < 0.05)", "Underexpressed (p < 0.05)"))
   theme_minimal() +
  labs(
    
    x = "Log2FoldChange",
    y = "-Log10(p-value)"
  )
pdf(file = paste0("plot_volcano", j, ".pdf"), width = 10, height = 6)
plot(volcano_)
dev.off()
}
}

```
-Gene Set Enrichment Analysis

```{r}

 
for (k in resultsNames(dds)){
if (k!="Intercept"){
 
  all_genes[[k]]$ress<- all_genes[[k]]$ress[order(-all_genes[[k]]$res$log2FoldChange),]
  genes<-rownames(all_genes[[k]]$res) #get gene esemble id
  
  genes_name<-mapIds(org.Hs.eg.db, keys = genes, column = "SYMBOL", keytype = "ENSEMBL") #map them to gene names
  all_genes[[k]]$ress$gene_names<- genes_name
  all_genes[[k]]$ress$gene_names[all_genes[[k]]$ress$gene_names == "<NA>"] <- NA
  all_genes[[k]]$ress<-all_genes[[k]]$ress[!is.na(all_genes[[k]]$ress$gene_names), ]#obtaining the genes that have a representative name

  all_genes[[k]]$ress <-all_genes[[k]]$ress[order(-all_genes[[k]]$ress$log2FoldChange),]
  all_genes[[k]]$ress <- all_genes[[k]]$ress %>%
  filter(padj <= 0.05)
 
  genes_list<- all_genes[[k]]$ress$log2FoldChange
  
  names(genes_list) <- rownames(all_genes[[k]]$ress)
  
  
  gse <- gseGO(genes_list,  #Gene Set Enrichment Analysis
             ont = "BP",
             keyType = "ENSEMBL",
             OrgDb = "org.Hs.eg.db",
             eps = 1e-300)
  all_genes[[k]]$gse<- as.data.frame(gse)
  all_genes[[k]]$gseplt<- gse
 # gseaplot(gse)
 # ridgeplot(gse)
 
  
}
}


```



```{r}
#Get known genes related to RNA STRESS GRANULES 
SG_genes <- read.csv("SG_genes.csv",header=F)
SG_names <- unlist(SG_genes$V1)

# for one condition: match the RNA SG with the pathways of the gene set enrichment results
common_SG <- intersect(SG_names, all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$ress$gene_names)
for (ind in seq(1:length(all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$gse$core_enrichment))){
    ensg_<-unlist(strsplit(all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$gse$core_enrichment[ind],"/"))
    enrch_gene_names<- all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$ress %>% filter(rownames(all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$ress) %in% ensg_) %>% select(gene_names)
    enrch_gene_names <- enrch_gene_names$gene_names
   

    
    all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$pathways[[all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$gse$Description[ind]]] <- intersect(enrch_gene_names,common_SG)
  
}
# for other condition
common_SG <- intersect(SG_names, all_genes$Condition_TDP43.wild.type_vs_Negative.control$ress$gene_names)
for (ind in seq(1:length(all_genes$Condition_TDP43.wild.type_vs_Negative.control$gse$core_enrichment))){
  
    ensg_<-unlist(strsplit(all_genes$Condition_TDP43.wild.type_vs_Negative.control$gse$core_enrichment[ind],"/"))
    enrch_gene_names<- all_genes$Condition_TDP43.wild.type_vs_Negative.control$ress %>% filter(rownames(all_genes$Condition_TDP43.wild.type_vs_Negative.control$ress) %in% ensg_) %>% select(gene_names)
    enrch_gene_names <- enrch_gene_names$gene_names
    

   
    all_genes$Condition_TDP43.wild.type_vs_Negative.control$pathways[[all_genes$Condition_TDP43.wild.type_vs_Negative.control$gse$Description[ind]]] <- intersect(enrch_gene_names,common_SG)
  
}





```






```{r}
 # Getting the Phantom files of the SG that belong in the  enriched pathways
folder_path <-"C:/Users/DELL/Desktop/RNA proj/hs.FANTOM.annotated"
files_in_folder <- list.files(folder_path)
genes_phantom<-c()
result_list <- list()
for (k in resultsNames(dds)){
  
  if (k!="Intercept" && length(all_genes[[k]]$gse!=8) ){
    selected_pathways <- names(Filter(function(x) length(x)>2, all_genes[[k]]$pathways))
      all_genes[[k]]$pathways<- all_genes[[k]]$pathways[selected_pathways] #filtering the keys
    unique_phantom_genes <- setdiff(unique(unlist( all_genes[[k]]$pathways)),genes_phantom) #SG genes in each condition uniquely
    common_SG_genes <- intersect(unique(unlist( all_genes[[k]]$pathways)), genes_phantom)

    
    genes_phantom <- unique(unlist(all_genes[[k]]$pathways))
    all_genes[[k]]$phantom <- unique_phantom_genes
    
    matching_files <- grep(paste("p1@(", paste(all_genes[[k]]$phantom, collapse = "|"),")(.*@.*)?\\.csv$", sep = ""), files_in_folder, value = TRUE)
    matching_files<-setdiff(matching_files,"T071118_p1@CALR3,p1@MED26.csv" )
    for (file in matching_files) { 
      data <- read.csv(paste0(folder_path, "/", file),skip=1)
      filtered_data <- data[data$Frel >= 0.98,] 
    result_list[[sub(".*@([^\\.]+)(.*@.*)?\\.csv$", "\\1",file)]]<- filtered_data
    }
  }
}
    ```



```{r}


#Filtering the gse objects that match rna stress granules
all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$gseplt@result <-all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$gseplt@result %>% filter(Description %in% names(all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$pathways))

all_genes$Condition_TDP43.wild.type_vs_Negative.control$gseplt@result <-all_genes$Condition_TDP43.wild.type_vs_Negative.control$gseplt@result %>% filter(Description %in% names(all_genes$Condition_TDP43.wild.type_vs_Negative.control$pathways))

```

# This part I got the leading edge of each enriched pathway that conatined RNA stress granules however none of the leading edge was inside the ranked genes of phantom files. 
```{r}
enrich_genes <- list()
for (ind in seq(1:length(all_genes$Condition_TDP43.wild.type_vs_Negative.control$gseplt@result$core_enrichment))){

ensg_<-unlist(strsplit(all_genes$Condition_TDP43.wild.type_vs_Negative.control$gseplt@result$core_enrichment[ind],"/"))

enrich_genes[[ind]] <-all_genes$Condition_TDP43.wild.type_vs_Negative.control$ress %>%  filter (row.names(.) %in% ensg_)

}
enrich_genes_ALL <- bind_rows(enrich_genes) %>%
    distinct()
SG_in_cond1 <- unlist(unique(all_genes$Condition_TDP43.wild.type_vs_Negative.control$pathways)) %>% unique()
values_to_remove <- c("RACK1", "YBX3", "SMN1", "HELZ2")
SG_in_cond1<- SG_in_cond1[!SG_in_cond1 %in% values_to_remove]

```

```{r}
#Obtaining enriched genes (leading edge) for first condition
enrich_genes <- list()
for (ind in seq(1:length(all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$gseplt@result$core_enrichment))){

ensg_<-unlist(strsplit(all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$gseplt@result$core_enrichment[ind],"/"))
enrich_genes[[ind]] <-all_genes$Condition_Asp_RRM2_Ala_vs_Negative.control$ress %>%  filter (row.names(.) %in% ensg_)

}


```






# Classifyig RNA stress granule genes found and checking which gene is common with other dataset
```{r}
#classifying the the Stress granules from gene enrichment as protein or RNA


SGs <-c("RPS19", "EIF3K", "RPS6", "EIF3B", "EIF3F", "EIF3D", "PABPC1", "RPS3", "RPS24", "EIF4B",
                 "EIF1", "EIF3H", "SLBP", "EIF3M", "EIF3I", "HNRNPD", "CNBP", "EIF4A1", "EIF3L", "EIF4H",
                 "CSDE1", "EIF2A", "EIF3E", "HABP4", "EIF4E3", "EIF3C", "RPS11", "CLNS1A", "EIF3G", "RAN",
                 "POP7", "SNRPF", "FBL", "RNASEL", "NCOA3", "MSI1", "MSI2", "FABP5", "EIF4G3", "CALR",
                 "SAMD4A", "TARDBP", "RBM47", "CNOT2")

# Create a dataframe with one column "sources"
SG_df <- data.frame(sources = SGs)

RNA_vs_Protein <- read.csv("C:/Users/DELL/Desktop/RNA proj/SG_RNA.csv")
SG_df$gene_type <- ifelse(SG_df$sources %in% RNA_vs_Protein$symbol, "RNA", "Protein")

```
```{r}
#Obtaining common SG from the two datasets
SG_from_other <- c('MKI67', 'CDC20', 'CDK1', 'CENPF', 'KIF23', 'SMC4', 'SPAG5', 'PHLDB2', 'ALPK2', 'DSP', 'TPM1', 'DTL', 'CDK2', 'BLM', 'PDLIM5', 'APOBEC3C', 'SERPINE1', 'ZFP36', 'TRIM56', 'PARP14', 'RBM47', 'YBX3', 'RBPMS', 'SOX3', 'ANXA1', 'PARK7')
intersect(SG_from_other,SGs)

```
